---
title: "R for Health Data Science"
subtitle: "Week 08: Biostats II"
author: "Sam Stewart"
date: "2021-04-09"
output: 
  pdf_document:
   number_sections: true
  html_document:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    # theme: cosmo
    # highlight: espresso
---
  
```{r setup, include=FALSE}

knitr::opts_chunk$set(tidy = TRUE, echo = TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5)

#This option should force knitr to use the project workng directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

options(full_width = FALSE)
options(knitr.kable.NA = '')

```

```{r readFram}
library(Hmisc)
library(kableExtra)
library(tidyverse)
library(ggplot2)
library(boomer)
library(survival)
library(survminer)
library(haven)
library(lme4)
library(mlmRev)

dat = read.csv("data/framinghamFirst.csv",header=TRUE, 
               na.strings=".",stringsAsFactors=FALSE)
dat$BMIGroups = cut(dat$BMI,breaks=c(0,18.5,25,30,Inf),
                    labels=c("Underweight","Normal","Overweight","Obese"))
dat$SEX = factor(dat$SEX,levels=1:2,labels=c("Male","Female"))
dat$DIABETES = factor(dat$DIABETES,levels=0:1,labels=c("No Diabetes","Diabetes"))
dat$HYPERTEN = factor(dat$HYPERTEN,levels=0:1,labels=c("Normotensive","Hypertensive"))

#for the survival analysis
dat.lung = lung %>% tibble %>%
  mutate(
    sex = factor(sex,levels=1:2,labels=c("M","F")),
    ageGroup = cut(age,breaks=c(0,50,60,70,100)),
    ph.ecog = factor(ph.ecog)
  )

#for the multilevel models
imm = read_dta("https://stats.idre.ucla.edu/stat/examples/imm/imm10.dta")

dat.imm = imm %>% 
  mutate(homework.fact = factor(homework))

```

This week we're going to get back to the Biostats/Epi world and cover some more complex modeling techniques, focusing on

1.  Survival Analysis
2.  Multilevel Models
3.  ~~Generalized Estimating Equations~~

# Survival Analysis
Survival analysis is the analysis of time to event data

* Time from surgery to infection
* Time from breast cancer diagnosis to death
* Time from entry into addiction treatment to relapse

We need three time points to perform a survival analysis

* Starting time (often 0)
* Stopping time
* Censoring - an indicator if the event occurred

We'll start with Kaplan Meier curves, then move onto Cox Proportional Hazards Regression.  We'll use a built-in R dataset here called `lung` on the survival rate of patients with advanced lung cancer. See `?lung` for more details.

## Kaplan-Meier
The command to fit a Kaplan-Meier curve is `survfit()`, and it uses the formula approach as with `lm()` and `glm()`. The difference here is that our outcome is a combination of either 2 or 3 variables - in this case there is a `time` variable that records the amount of time followed, and a `status` variable that records whether whether the subject died or not.  We'll pass those to a `Surv()` object on the left side of the equation, and then the stratifying variables are on the right.


```{r km01}

km.lung00 = survfit(Surv(time,status)~1,data=dat.lung)
km.lung00
summary(km.lung00)

km.lung01 = survfit(Surv(time,status)~sex,data=dat.lung)
km.lung01
summary(km.lung01)

```

The printout of the model reports the basic summary stats, and the summary reports the lifetable.  We can also plot the data, either with the basic `plot()` function, or with the ggplot-friendly `survminer::ggsurvplot()`.  `ggsurvplot()` doesn't follow the `tidyverse` principles perfectly, but it's still a marked improvement from the base library, see [this blog](https://rpkgs.datanovia.com/survminer/) for a good exploration of the options.

```{r km02}

plot(km.lung00)
plot(km.lung01,col=2:3,xlab='Days',ylab='Survival Probabilty',main='Lung cancer survival rate by Sex')
legend("topright",fill=2:3,legend=c("Male","Female"))
```

```{r km03,fig.height=8,fig.width=8}
ggsurvplot(km.lung00)

ggsurvplot(km.lung01,
           ggtheme=theme_grey(),
           conf.int=TRUE,
           pval=TRUE,
           risk.table=TRUE)+
  labs(x="Days",Strata="")

```

## Cox PH Regression
We'll use the `coxph()` command to fit Cox PH regression models - they follow the same general format as `survfit()`, but produce a regression model.  We'll evaluate them using the `car::Anova()` command, which will evaluate the varialbes using log-likelihood tests.

```{r coxph01}

#simple model predicting survival with sex
mod.lung01 = coxph(Surv(time,status)~sex,data=dat.lung)
#Anova evaluates each variables
car::Anova(mod.lung01)
#getting the coefficient table
summary(mod.lung01)

#full model predicting survival with sex, ageGroup and ECOG score
mod.lung02 = coxph(Surv(time,status)~sex+ageGroup+ph.ecog,data=dat.lung)
car::Anova(mod.lung02)
summary(mod.lung02)

```

That's good, but we need to get this in production-quality, so we'll format using our `tidyverse` skills.

```{r coxph02}

#in multiple steps like a sane person
#a) extract the coefficients
coef = summary(mod.lung02)$coefficients %>% as.data.frame() %>%
  rownames_to_column()
#b) extract the confidence intervals
ci = summary(mod.lung02)$conf.int %>% as.data.frame %>%
              rownames_to_column()
#c) join the two, and reduce to useful columns
out01 = left_join(coef,ci) %>% 
  select(rowname,'exp(coef)','lower .95','upper .95','Pr(>|z|)') %>%
  rename(RR=`exp(coef)`,
         variable = rowname)
#d) print the columns
out01 %>% kable(digits=c(0,3,2,2,4)) %>% kable_styling()

#OR
#do it all at once
summary(mod.lung02)$coefficients %>% as.data.frame %>% 
  rownames_to_column() %>%
  left_join(summary(mod.lung02)$conf.int %>% as.data.frame %>%
              rownames_to_column()
              ) %>%
  select(rowname,'exp(coef)','lower .95','upper .95','Pr(>|z|)') %>%
  rename(RR=`exp(coef)`,
         variable = rowname) %>% 
  kable(digits=c(0,3,2,2,4)) %>% kable_styling() %>%
  pack_rows("Sex",1,1)%>%
  pack_rows("Age Group",2,4)%>%
  pack_rows("ECOG Score",5,7)

```

# Multilevel Models
There are several ways to fit multilevel models in R, but the most common are with the `lmer()` and `glmer()` functions from the `lme4` library.  I'll focus on their use for multilevel models with random intercepts and/or random slopes - for longitudinal data I prefer `gee()`, but if you like mixed-effects for longitudinal data you can use `lmer()`.

For this section I'm going to follow along with the UCLA Stats page on multilevel modeling, available [here](https://stats.idre.ucla.edu/r/examples/mlm-imm/r-kreft-chp-3/).

# Syntax
The traditional syntax is `y~x` for a simple model, or `y~x1+x2+...+xk` for a multiple model.  With multilevel modeling we want to be able to adjust for random slopes and/or random intercepts, which we control with additional random components in brackets `()` at the end of the equation. For a **random intercept** the format will be 


```
y ~ x1+x2+x3+(1|group)
```

The `1` in the random component refers to the intercept, in a similar fashion to a regression model `y~1` that fits y predicted with only an intercept. This fits a model that predicts `y` with `x1, x2, x3`, but adds a random intercept for each level of `group`.

If we also want random slopes the format would be

```
y ~ x1+x2+x3+(x2|group)

```

In this equation we're adding a random slope for the variable `x2`, while also including a random intercept by default.  Note that `x2` doesn't need to be in both variable lists, that was our choice for this model.

## `lmer` example from UCLA stats
I'm having a hard time finding the provenance of the `imm10` dataset - it's from the UCLA Stats textbook *Introduction to Multilevel Modeling*, but I can't find a help file or description of it, so I'm going to infer most of it.

It has 260 subjects from 10 different schools - we're going to focus on their math scores as a function of their dedication to homework.


```{r imm10.explore}

dat.imm %>%
  ggplot(aes(x=math))+
  geom_histogram()

#geom_jitter is like geom_point, but can scatter things a bit randomly, useful for overlapping datapoints
dat.imm %>%
  ggplot(aes(x=homework,y=math))+
  geom_violin(aes(group=homework,colour=homework.fact))+
#  geom_point(aes(colour=homework.fact))+
  geom_jitter(aes(colour=homework.fact),width=0.2,height=0)+
  geom_smooth()

dat.imm %>%
  ggplot(aes(x=homework,y=math))+
  geom_jitter(width=0.2,height=0)+
  geom_smooth()
  
dat.imm %>%
  ggplot(aes(x=math))+
  geom_histogram()+
  facet_wrap(~schid)

dat.imm %>%
  ggplot(aes(x=homework,y=math))+
  geom_point(aes(colour=homework.fact))+
  geom_smooth()+
  facet_wrap(~schid)

```

The graphics demonstrate a relationship between math score and homework, but probably significant variation between schools.  We'll start with individual linear regression models.

```{r imm.linear01}

lin.mods = by(dat.imm,dat.imm$schid,function(x){lm(math~homework,data=x)})

```

The `by` function stratifies a dataset by a variable and performs a function on each stratified dataset - in this case I fit a simple linear regression predict math score with homework for each school individually.  I want to see how the models differ.

```{r imm.linear02}

t(sapply(lin.mods,coef))

```

These numbers confirm what our last plot showed - that the relationship between math and homework does not seem to be consistent across schools.

Based on the differences in both intercept and slope values, we'll fit a mixed effects model with a random slope for homework per school (and a random intercept per school).

```{r imm.lme01}

mod.lme01 = lmer(math~homework+(homework|schid),data=dat.imm)
mod.lme01
car::Anova(mod.lme01)
summary(mod.lme01)

```

And we can expand the model to other fixed effects as well (including school-level effects)

```{r imm.lme02}

mod.lme02 = lmer(math~homework+public+(homework|schid),data=dat.imm)
mod.lme02
car::Anova(mod.lme02)
summary(mod.lme02)

mod.lme03 = lmer(math~homework*public+(homework|schid),data=dat.imm)
mod.lme03
car::Anova(mod.lme03)
summary(mod.lme03)

```

# Breakout Activity
1.  There's another built-in dataset called `veteran` that reports the results of a trial comparing two treatment regimens for people with lung cancer.  Use the dataset to
    + Draw a Kaplan-Meier curve survival with treatment
    + Fit a Cox PH model predicting survival with treatment, cancer type and age
2.  For mixed effects model there's a library called `mlmRev` that has a dataset called `Exam` - it records the exam scores for 4059 students from 65 schools in Inner London.  I want you to build a model to predict their normalized exam score (`normexam`) based on the available data, see `?Exam` after you install and load the library.