---
title: "Week 05 - Breakout Activity"
author: "Sam Stewart"
date: "2021-03-19"
output: 
  # pdf_document:
  #  number_sections: true
  html_document:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
---
  
```{r setup, include=FALSE}

library(Hmisc)
library(kableExtra)
library(dplyr)
library(tidyr)
library(epiR)

#Note that I changed echo=FALSE - normally we don't want to include our code
knitr::opts_chunk$set(tidy = TRUE, echo = FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5)

#this is specific to my R-project code to find the data set correctly in the next block, you can probably delete it
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

#these are table-specific options
options(full_width = FALSE)
options(knitr.kable.NA = '')

```

```{r readFram}

dat = read.csv("data/framinghamFirst.csv",header=TRUE, 
               na.strings=".",stringsAsFactors=FALSE)
dat$BMIGroups = cut(dat$BMI,breaks=c(0,18.5,25,30,Inf),
                    labels=c("Underweight","Normal","Overweight","Obese"))
dat$SEX = factor(dat$SEX,levels=1:2,labels=c("Male","Female"))
dat$DIABETES = factor(dat$DIABETES,levels=0:1,labels=c("No Diabetes","Diabetes"))
dat$HYPERTEN = factor(dat$HYPERTEN,levels=0:1,labels=c("Normotensive","Hypertensive"))
dat$CURSMOKE = factor(dat$CURSMOKE,levels=0:1,labels=c("Non-smoker","Smoker"))

```

# Introduction
In week 4 we were presented with 4 questions (dropped the 5th for time):

1. Is there a difference in glucose levels between smokers and non-smokers?
2. Is the difference still present after controlling for sex?
3. Is there an interaction between smoking and sex?
4. Is there an effect of smoking on *hypertension*?

*NOTE: I changed diabetes for hypertension from last week - I always forget how much lower the diabetes rate was in the 80s: [https://www.cdc.gov/diabetes/statistics/slides/long_term_trends.pdf](https://www.cdc.gov/diabetes/statistics/slides/long_term_trends.pdf).* 

I'll split these into two sections, and explore the answers

# Glucose and Smoking
The primary question is if there is a difference in glucose between smokers and non-smokers.  We'll start with a tabular and graphical summary of the relationship.
```{r glucoseBoxplot, fig.width=8,fig.height=5}

par(mfrow=c(1,2))
boxplot(GLUCOSE~CURSMOKE,data=dat,col=c(3,2),xlab='',ylab='Glucose')
boxplot(GLUCOSE~CURSMOKE,data=dat,col=c(3,2),xlab='',ylab='Glucose (truncated)',ylim=c(40,150))

```

The boxplots show a minimal difference and perhaps some skewness, let's look at a tabular summary.

```{r glucoseTable}

#This is the equivalent of the STATA command
#by SEX: codebook GLUCOSE

temp0 = psych::describe(dat$GLUCOSE,quant = c(0.25,0.75))
temp1 = psych::describe(dat$GLUCOSE[which(dat$CURSMOKE=='Smoker')],quant = c(0.25,0.75))
temp2 = psych::describe(dat$GLUCOSE[which(dat$CURSMOKE=='Non-smoker')],quant = c(0.25,0.75))
#I have to wrap the rbind in an as.data.frame to make the subsetting command work
tab.sex01 = as.data.frame(rbind(
  ALL=temp0,
  Smoker=temp1,
  `Non-smoker`=temp2
))
tab.gluc.smoke = tab.sex01[,c('n','mean','sd','min','Q0.25','median','Q0.75','max')]

kable_styling(kable(tab.gluc.smoke,digits=2,caption='Glucose summarized across smoking status (manually, using describe)'))

#there's a better way using a library called table1
library(table1)
table1(~GLUCOSE|CURSMOKE,data=dat,caption='Glucose levels, overall and stratified by smoking status (automatically, using table1)')
```

```{r glucoseTest}
#the actual test would be a good idea
test.gluc.smoke = t.test(GLUCOSE~CURSMOKE,data=dat)
test.gluc.smoke
```

The difference looks statistically significant, though whether it is clinically meaningful is a different discussion.

**OBJECTIVE 1** Summarize the t-test in a meaningful way, either with a table or in-line.

## Investigating relationship to SEX
We want to investigate the effect of sex on this relationship, first as a controlling factor and second as an effect modifier.

The table below presents the stratified effects, and probably suggests a lack of effect modification, but we'll pursue both models regardless.
```{r gluc.sex.table}
#table 1 can be useful as a first step here
table1(~GLUCOSE|SEX+CURSMOKE,data=dat,caption='Glucose levels, stratified by Sex and Smoking Status')
```

For the regression models we'll build three - one with smoking, one with smoking and sex, one with their interaction also included.
```{r gluc.sex.models}
#obviously the regression model is most appropriate
mod01 = lm(GLUCOSE~CURSMOKE,data=dat)
summary(mod01)
mod02 = lm(GLUCOSE~CURSMOKE+SEX,data=dat)
summary(mod02)
mod03 = lm(GLUCOSE~CURSMOKE*SEX,data=dat)
summary(mod03)

```

The effect of sex, both as a controlling and effect-modifying variable, is minimal.  

**OBJECTIVE 2** Summarize the regression models in a more effective way.  If you're feeling really ambitious, look up the `stargazer` library for an effective way to summarize multiple models at once.

# Hypertension and Smoking
We'll start with a simple graphical and numeric summary of the relationship

```{r hyper.smoke,fig.height=5,fig.width=5}

tab.hyper.smoke = table(dat$CURSMOKE,dat$HYPERTEN)[2:1,2:1]
barplot(tab.hyper.smoke,beside=TRUE,legend=rev(levels(dat$CURSMOKE)),col=c(2,4))

table1(~HYPERTEN|CURSMOKE,data=dat)
```
The table suggests a pattern - to test and measure the effect we'll use the RR, OR and RD, as reported in the output below.

```{r hyper.smoke.test}
#getting the effect measures
test.hyper.smoke = epi.2by2(tab.hyper.smoke)
test.hyper.smoke
```

All three effects are significant, suggesting a protective effect of smoking on hypertension. 

**OBJECTIVE 3** Report the effect measures (RR, OR, RD) with their CIs in an effective way.